---
title: "A short report on the **testReads** file"
author: "Khristian Kotov"
date: "25.02.2015"
output: pdf_document
---
# Prerequisites

Input file [testReads.fastq](http://dl.dropbox.com/u/68829208/testReads.fastq.gz)

1. This file contains NGS reads for some human DNA regions.
2. The fastq file format is described at http://en.wikipedia.org/wiki/FASTQ_format, the
Phred quality score is in sanger format not illumina.
3. The report should be in PDF format and figures rather than descriptive words is preferred. Please also make clear description on your figures.
4. The report could be anything that you find interesting on the file. For example, it could be patterns you found on the data, com

# Exploratory analysis of the headers

Each record header consists of five fields:
*unique id*, *rank*, *x*, *y*, and *length*. An example looks as follows:

@HKVCQHG01APQJG rank=0000016 x=176.0 y=10.0 length=108

The basic distributions of these variables do not seem to tell us any particularly interesting story:

```{r, echo=F, message=FALSE, fig.height=4}

if( !file.exists("testReads.fastq") ){
   print("Downloading original data")
   download.file("http://dl.dropbox.com/u/68829208/testReads.fastq.gz", destfile="testReads.fastq.gz")
   ungzip("testReads.fastq.gz")
}

if( !file.exists("qwe.csv") ){
   system("grep '^@' testReads.fastq | sed -e 's|=| |g' -e 's|rank 0*|rank |g' -n -e 's|^@HKVCQHG01||p' | awk 'BEGIN {print \"ID,rank,x,y,len\"} {print $1\",\"$3\",\"$5\",\"$7\",\"$9}' > qwe.csv")
}

library(ggplot2)
qwe <- read.csv("qwe.csv",head=T,sep=',')

p1 <-ggplot(data=qwe,aes(x=rank)) + geom_histogram(fill="steelblue") + scale_x_continuous(breaks=seq(0,240000,60000))
p2 <-ggplot(data=qwe,aes(x=x)) + geom_histogram(fill="steelblue")
p3 <-ggplot(data=qwe,aes(x=y)) + geom_histogram(fill="steelblue")
p4 <- ggplot(data=qwe,aes(x=len)) + geom_histogram(breaks=seq(0,1000,1),fill="steelblue")

library(gridExtra)
grid.arrange(p1, p4, p2, p3, ncol=2)
```

Neitgher do correlations tell us an interesting story. So let us omit rather non-informative scatter plots here.
<!-- (except for a weak positive *rank* <-> *length*) -->


# Analysis of the sequences and their qualities

Objectives:

- identify data variability
- find repeating patterns and do a pattern density check
- study the data quality
- look for various correlations

Analysis design:

- pick a pattern length from 1 to 32 symbols
- scan sequences and count all unique patters
- identify patters that repeat more often than others
- calculate mean and standard deviation of qualities for these pattern
- evaluate probability of the pattern to be read correctly

Starting from the code that I already developed for the task #1, I implemented [a simple command line tool](https://github.com/koskot77/sophia/blob/master/analysis.cc) that runs a density check for patterns of a specified length. In the trivial case of *length=1* we have four patters after the four bases:

```{r, echo=F, message=FALSE}

if( !file.exists("sophia/") ){
   system("git clone https://github.com/koskot77/sophia.git")
   system("cd sophia && g++ -Wall -o test analysis.cc && for i in 1 2 3 4 5 6 7 8 9 10 11 12 13 14 15 16 17 18 19 20 21 22 23 24 25 26 27 28 29 30 31 32 ; do time ./test $i ; mv output.csv output_$i.csv ; done")
}
an <- read.csv(file='sophia/output_1.csv',head=T,sep=',')
print(an)
```

Number of possible combinations grows as 4^*length* until *length*~10 where it slows down (patterns become less random while still growing exponentially in numbers):

```{r, echo=F, message=FALSE, fig.height=4}
#combinations <- data.frame( width=integer(), found=integer() )
#for(w in 1:32){
#  print(w)
#  an<-read.csv(file=paste('sophia/output_',w,'.csv',sep=""),header=T,sep=',')
#  combinations <- rbind( combinations, data.frame( width=w, found=log2(dim(an)[[1]]) ) )
#}
#print(combinations)
combinations <- data.frame( width=seq(1,32,1), found=c(
2, 4, 6, 8, 10, 12, 13.99, 15.89, 17.36, 18.24, 18.71, 18.97,
19.13, 19.25, 19.36, 19.46, 19.55, 19.63, 19.72, 19.79, 19.87, 19.94,
20.01, 20.07, 20.14, 20.20, 20.26, 20.32, 20.38, 20.43, 20.48, 20.50) )

ggplot(data=combinations,aes(x=width,y=found)) + geom_point() + labs(x="pattern length (symbols)",y="log2( # of unique patterns found )",title="Data variability")
```

Let us check how often the patterns repeat in the data:

```{r, echo=F, message=FALSE, fig.height=4}
an30 <- read.csv(file='sophia/output_30.csv',header=T,sep=',')
p1 <- ggplot(data=an30,aes(x=ID,y=count)) + geom_point(color="steelblue",size=0.5) + labs(x="pattern ID",y="# of patterns found",title="Pattern length = 30")

an20 <- read.csv(file='sophia/output_20.csv',header=T,sep=',')
p2 <- ggplot(data=an20,aes(x=ID,y=count)) + geom_point(color="steelblue",size=0.5) + labs(x="pattern ID",y="# of patterns found",title="Pattern length = 20")

an15 <- read.csv(file='sophia/output_15.csv',header=T,sep=',')
p3 <- ggplot(data=an15,aes(x=ID,y=count)) + geom_point(color="steelblue",size=0.5) + labs(x="pattern ID",y="# of patterns found",title="Pattern length = 15")

an10 <- read.csv(file='sophia/output_10.csv',header=T,sep=',')
p4 <- ggplot(data=an10,aes(x=ID,y=count)) + geom_point(color="steelblue",size=0.5) + labs(x="pattern ID",y="# of patterns found",title="Pattern length = 10")

grid.arrange(p4, p3, p2, p1, ncol=2)
```

Interesting features on these plots are:

1. "vertical lines" around some of the pattern IDs (e.g. see 
"length = 30" plot)
2. "horisintal lines"

Close look into the patterns from the structures confirmed that they often correspond to some variations of the same sequences. In other words, these patterns get assigned with the different IDs, but they often share parts.

Let us now look at the adjusted to 0 (shifted by 33 units) Phred quality scores of the frequent patterns:

```{r, echo=F, message=FALSE, fig.height=4}

p14 <- ggplot(data=an10[an10$count>30000,],aes(x=meanQual,y=count)) + geom_point(color="steelblue",size=1) + labs(x=expression(paste("Average quality (",Sigma," q / length)")),y="# of patterns",title="Pattern length = 10")

p13 <- ggplot(data=an15[an15$count>10000,],aes(x=meanQual,y=count)) + geom_point(color="steelblue",size=1) + labs(x=expression(paste("Average quality (",Sigma," q / length)")),y="# of patterns",title="Pattern length = 15")

p12 <- ggplot(data=an20[an20$count>5000,],aes(x=meanQual,y=count)) + geom_point(color="steelblue",size=1) + labs(x=expression(paste("Average quality (",Sigma," q / length)")),y="# of patterns",title="Pattern length = 20")

p11 <- ggplot(data=an30[an30$count>1000,],aes(x=meanQual,y=count)) + geom_point(color="steelblue",size=1) + labs(x=expression(paste("Average quality (",Sigma," q / length)")),y="# of patterns",title="Pattern length = 30")

grid.arrange(p14, p13, p12, p11, ncol=2)
```



# Summary

In addition, 432 records with base 'N' and quality 0 were identified. Patterns that included this base were dismissed from the study.
